---
title: "2.3 Conditional Statements - Tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)

library(ottr)
library(testthat)
```

---

At the end of this tutorial you should be able to:

*Implement if statements
*Implement if-else statements
*Implement if-else if-else statements
*Nest conditional statements within one another
*Apply conditional statements to data

---

## 2.3.1: If & If-else Statements

---

“The IMPC is an international effort by 21 research institutions to identify the function of every protein-coding gene in the mouse genome.” (https://www.mousephenotype.org) To achieve this the IMPC knocks out each mouse gene individually and uses a panel of hundreds of standardized tests to assess mouse phenotypes, this includes, but is not limited to: assessment of behavioural tendencies, x-rays to examine the skeletal system, blood tests, and observations of morphological differences. These can be compared to control to determine which phenotypes are affected by the knocked out gene.

In this tutorial we will work with data for 4 mice with the Ap4e1 gene knocked out (mutant mice), and data for 4 control mice. This table shows the results of a test to assess the hemoglobin content (in g/dL) along with an ID number, mouse sex, and group (mutant or control). 
(Data from: https://www.mousephenotype.org)

| ID # | Sex    | Group   | Hemoglobin (g/dL) |
| ---- | ------ | ------- | ----------------- |
| 1    | female | mutant  | 14.9              |
| 2    | female | mutant  | 15.1              |
| 3    | male   | mutant  | 14.6              |
| 4    | male   | mutant  | 15.2              |
| 5    | female | control | 13.4              |
| 6    | female | control | 13.9              |
| 7    | male   | control | 13.4              |
| 8    | male   | control | 14.9              |

Below the variables you will work with in this tutorial are defined. The sex of each mouse is saved as "mouse_ID#_sex", the group as "mouse_ID#_group", and the hemoglobin content as "mouse_ID#_hemo".

```{r}
mouse_1_sex <- "female"
mouse_1_group <- "mutant"
mouse_1_hemo <- 14.9
mouse_2_sex <- "female"
mouse_2_group <- "mutant"
mouse_2_hemo <- 15.1
mouse_3_sex <- "male"
mouse_3_group <- "mutant"
mouse_3_hemo <- 14.6
mouse_4_sex <- "male"
mouse_4_group <- "mutant"
mouse_4_hemo <- 15.2
mouse_5_sex <- "female"
mouse_5_group <- "control"
mouse_5_hemo <- 13.4
mouse_6_sex <- "female"
mouse_6_group <- "control"
mouse_6_hemo <- 13.9
mouse_7_sex <- "male"
mouse_7_group <- "control"
mouse_7_hemo <- 13.4
mouse_8_sex <- "male"
mouse_8_group <- "control"
mouse_8_hemo <- 14.9
```

Write an if statement to check if mouse 1 is female and if so, prints "Mouse is female". Do not alter the line containing the code `q.1.a <-` or the otter grader will not be able to check your answer.
```{r}
q.1.a <-
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.1.a.R")
```
**Check your answer by running the above block** 

Write an if statement to check if mouse 7 has a hemoglobin content greater than 15 g/dL and if so, assigns the value "high" to a variable named "hemo_content".
```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.1.b.R")
```
**Check your answer by running the above block** 

Write an if statement to check if mouse 4 is male AND is a mutant mouse, if so print the hemoglobin content of the mouse. Do not alter the line containing the code `q.1.c <-` or the otter grader will not be able to check your answer.
```{r}
q.1.c <-
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.1.c.R")
```
**Check your answer by running the above block** 

Write an if-else statement that checks if mouse 6 is a mutant. If the mouse is a mutant, print "Mutant mouse", if it is not a mutant, print "Control mouse". Do not alter the line containing the code `q.1.d <-` or the otter grader will not be able to check your answer.
```{r}
q.1.d <-
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.1.d.R")
```
**Check your answer by running the above block** 

Write an if-else statement that checks if mouse 2 is a male. If the mouse is a male, assign the value "M" to a variable named "sex", if it is not, assign the value "F" to a variable named "sex".
```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.1.e.R")
```
**Check your answer by running the above block** 

The mean hemoglobin content is 14.95 for mutant mice and 13.9 for control mice.

Write an if-else statement that will:
- Assign the absolute value of the difference between the mean MUTANT hemoglobin content and the hemoglobin content of mouse 5 to a variable called "diff.from.group.mean", IF the mouse is a MUTANT mouse.
- Assign the absolute value of the difference between the mean CONTROL hemoglobin content and the hemoglobin content of mouse 5 to a variable called "diff.from.group.mean", IF the mouse is a CONTROL mouse.
```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.1.f.R")
```
**Check your answer by running the above block** 

---

## 2.3.2: If-else if-else Statements

---

Write an if-else-if-else statement that does the following:
If mouse 4 has a hemoglobin content < 14 g/dl, print "Hemoglobin < 14 g/dL".
Else, if mouse 4 has a hemoglobin content >= 14 g/dl and < 15 g/dl, print "Hemoglobin >= 14 g/dL and < 15 g/dL". 
Else, print "Hemoglobin >= 15 g/dL."

Write the statement WITHOUT using a logical expression containing `&`, and WITHOUT USING the comparison operators `>` and `>=`. Do not alter the line containing the code `q.2.a <-` or the otter grader will not be able to check your answer.
```{r}
q.2.a <-
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.2.a.R")
```
**Check your answer by running the above block** 

Why don't you need to use the logical expression `mouse_4_hemo >= 14 & mouse_4_hemo < 15`?

There are 4 possible combinations of mouse sex and group (4 mutually exclusive possibilities!):

- female mutant
- male mutant
- female control
- male control

Write an if-else if-else statement that will check which of these 4 possibilities fits mouse 7 and saves it to a variable called "category". For example, if mouse 7 is a female mutant, `category <- "female mutant"`, etc. 
```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.2.b.R")
```
**Check your answer by running the above block** 

Examine the following if-else if-else statement.

```{r}
if (mouse_3_sex == "male") {
  print("Male mouse")
} else if (mouse_3_sex == "female") {
  print("Female mouse")
} else if (mouse_3_group == "mutant") {
  print("Mutant mouse")
} else {
  print("Control mouse")
}
```

Answer the following questions:

- Which logical expressions would return `TRUE`?
- Which logical expression would return `TRUE` but the following code chunk is not executed? Why isn't it executed?
- Which print statements will never be run no matter which mouse you use?
- After answering all of these questions: what is wrong with this if-else if-else statement?

---

## 2.3.3: Nested Conditional Statements

---

Write an if statement nested within another if statement. If mouse 5 is female, check if mouse 5 is a control mouse, and then print "Mouse is a female control". Do not alter the line containing the code `q.3.a <-` or the otter grader will not be able to check your answer.
```{r}
q.3.a <-
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.3.a.R")
```
**Check your answer by running the above block** 

Write an if-else if-else statement nested within an if statement. The if statement should check if mouse 1 is a mutant, if it is, then an if-else if-else statement should check if the mouse has a hemoglobin content:
- less than 14 (if so, print "hemoglobin < 14")
- greater than or equal to 14 and less than 15 (if so, print "14 <= hemoglobin < 15")
- or greater than 15 (if so, print "hemoglobin >= 15")
If the mouse is not a mutant, nothing should be printed.

Do not alter the line containing the code `q.3.b <-` or the otter grader will not be able to check your answer.
```{r}
q.3.b <-
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.3.b.R")
```
**Check your answer by running the above block** 

Use one or more conditional statements to write code that does the following:
- Checks if mouse 8 is a mutant mouse or a control mouse, and calculates the absolute value of the difference from the group's mean hemoglobin content. This value should be assigned to a variable named "diff.from.group.mean" 
- If the absolute value of the difference is > 1 standard deviation away from the group mean, the value `TRUE` is assigned to a variable named "over.1.sd", otherwise the value `FALSE` is assigned to a variable named "over.1.sd".

Necessary values for calculations:

| group   | mean  | standard deviation |
| ------- | ----- | ------------------ |
| mutant  | 14.95 | 0.26               |
| control | 13.9  | 0.71               |
```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W2.3.3.c.R")
```
**Check your answer by running the above block** 

---