---
title: "Week 2 Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)

library(ottr)
library(testthat)
```

---

## Overview

---

This assignment covers the content in week 2 lectures and tutorials. 

#### Mark breakdown
- Question 1 (5 marks)
- Question 2 (10 marks)
- Question 3 (8 marks)

---

## Background

---

Data in this assignment are based on information from:

The ENCODE Project Consortium., Moore, J.E., Purcaro, M.J. et al. Expanded encyclopaedias of DNA elements in the human and mouse genomes. Nature 583, 699–710 (2020).

Promoter and enhancer regions regulate the transcription of genes.

- A *promoter region* is a sequence just upstream of the transcription start site (TSS) of a gene that enables transcription initiation. 

- An *enhancer region* is a sequence that influences the rate of transcription of a particular gene. It does not have a specific location with respect to the gene it regulates, although they are cis-acting (on the same chromosome). A *proximal enhancer* is an enhancer that is close the TSS.

Promoters and enhancers have the following characteristics:

|                   | Promoter region | Enhancer region      |
| ----------------- | --------------- | -------------------- |
| Distance from TSS | < 200 bp        | no specific distance |
| DHS               | TRUE            | TRUE                 |
| H3K4me3 level     | high            | low                  |
| H3K27ac level     | low             | high                 |


- *DHS*: active promoter and enhancer regions contain DNAse I hypersensitivity sites (DHSs). In these data, regions that contain DHSs will have a "DHS" value of TRUE, indicating that the regions could be a promoter or an enhancer.

- *H3K4me3*: a histone mark that is present at a HIGH level in promoter regions and a LOW level in enhancer regions

- *H3K27ac*: a histone mark that is present at a LOW level in promoter regions and a HIGH level in enhancer regions

We will start by examining three regions in the human genome: A, B, and C. Characteristics for these three regions will be assigned to the following variables (X represents A, B, or C):

- `rX_dist_TSS`     distance from closest TSS (numeric)
- `rX_DHS`          is it a DNAse I hypersensitivity site (logical)
- `rX_H3K4me3`      level of H3K4me3: high or low (character)
- `rX_H3K27ac`      level of H3K27ac: high or low (character)

```{r}
# values for region A
rA_dist_TSS <- 158
rA_DHS <- TRUE
rA_H3K4me3 <- "high"
rA_H3K27ac <- "low"

# values for region B
rB_dist_TSS <- 192
rB_DHS <- FALSE
rB_H3K4me3 <- "low"
rB_H3K27ac <- "low"

# values for region C
rC_dist_TSS <- 4729
rC_DHS <- TRUE
rC_H3K4me3 <- "low"
rC_H3K27ac <- "high"
```

---

## QUESTIONS 

---

### Question 1

#### a. 
(1 mark)
Region A is closest to the TSS of the gene "MBNL1". Convert the variable `rA_gene_lowercase` defined below to uppercase letters using one of the functions we learned this week. Assign the value to a variable named "q.1.a".

```{r Assignment.W2.1.a}
rA_gene_lowercase <- "mbnl1"
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.1.a.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### b.
(1 mark)
Using one of the functions we learned this week, get the number of letters in the value saved to variable `rA_H3K27ac` (i.e., not in the variable name "rA_H3K27ac" itself). Assign the result to a variable named "q.1.b".

```{r Assignment.W2.1.b}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.1.b.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### c. 
(1 mark)
The mean distance of regions A, B, and C from the TSS that is closest to each of them is calculated below and assigned to the variable `mean_dist_TSS`. Using one of the functions we learned this week, round `mean_dist_TSS` to two digits after the decimal place. Assign the return value to a variable named "q.1.c".

```{r Assignment.W2.1.c}
mean_dist_TSS <- (rA_dist_TSS + rB_dist_TSS + rC_dist_TSS)/3
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.1.c.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### d.
(2 marks)
By nesting two functions that we learned this week, extract the 2nd and 3rd characters of the value saved to variable `rC_H3K4me3` and output a logical value that indicates whether those 2 letters are also found in the value saved to variable `rA_H3K27ac` (i.e., do not compare the variable names). Assign the return value to a variable named "q.1.d".

```{r Assignment.W2.1.d}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.1.d.R")
```
**IGNORE the above block: test results will not appear in assignments** 

---

### Question 2

#### a.
(1 mark)
Write a logical expression that answers the following question: Is region B close enough to a TSS for region B to be a promoter? 
Remember: promoters must be within 200 bp of the TSS.
Assign the return value of the logical expression to a variable named "q.2.a".

```{r Assignment.W2.2.a}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.2.a.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### b.
(1 mark)
Write a logical expression that answers the following question: Is the level of H3K27ac the same in region B as it is region C? 
Assign the return value of the logical expression to a variable named "q.2.b".

```{r Assignment.W2.2.b}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.2.b.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### c.
(1 mark)
Write a logical expression that answers the following question: Is the distance of region C from the closest TSS greater than or equal to the distance from the closest TSS for region B? 
Assign the return value of the logical expression to a variable named "q.2.c".

```{r Assignment.W2.2.c}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.2.c.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### d.
(1 mark)
Write a logical expression that answers the following question: Is region A close enough to a TSS for region A to be a promoter and is region B close enough to a TSS for region B to be a promoter?
Assign the return value of the logical expression to a variable named "q.2.d".

```{r Assignment.W2.2.d}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.2.d.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### e. 
(1 mark)
Write a logical expression that answers the following question: Do region B or region C have a high level of H3K27ac?
Assign the return value of the logical expression to a variable named "q.2.e".

```{r Assignment.W2.2.e}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.2.e.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### f. 
(1 mark)
Write a logical expression that answers the following question: Does region C NOT have a high level of H3K4me3 AND does it NOT have low level of H3K4me3?
Assign the return value of the logical expression to a variable named "q.2.f".

```{r Assignment.W2.2.f}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.2.f.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### g. 
(2 marks)
Write a logical expression that answers the following question: Do any of the regions (A, B, or C) NOT contain DNAse hypersentivity sites?
Assign the return value of the logical expression to a variable named "q.2.g".

```{r Assignment.W2.2.g}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.2.g.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### h. 
(2 marks)
Write a logical expression that answers the following question: Does region A or region B contain DNAse hypersentivity sites AND does region A or region B have a high level of H3K27ac?
Assign the return value of the logical expression to a variable named "q.2.h".

```{r Assignment.W2.2.h}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.2.h.R")
```
**IGNORE the above block: test results will not appear in assignments** 

---

### Question 3

#### a. 
(1 mark) 
Write an if statement that checks if region C contains DNAse hypersensitivity sites and if it does, prints "Region is likely a promoter or an enhancer". Do not alter the line containing the code `q.3.a <-` or you will not receive a mark for this question.

```{r Assignment.W2.3.a}
q.3.a <- # Do not modify this line
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.3.a.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### b. 
(1 mark)
Write an if-else statement that checks if region B is within 200 base pairs of a TSS. If it is within 200 base pairs print "Region is potentially a promoter", otherwise print "Region cannot be a promoter". Do not alter the line containing the code `q.3.b <-` or you will not receive a mark for this question.

```{r Assignment.W2.3.b}
q.3.b <- # Do not modify this line
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.3.b.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### c.
(3 marks)
Use one or more conditional statements to write code that checks the values of the 4 arguments defined below (`rX_DHS`, `rX_dist_TSS`, `rX_H3K4me3`, `rX_H3K27ac`) to determine if the region is likely a promoter, likely an enhancer, or neither. 

- If the variables match the criteria for a promoter, the value "promoter" should be assigned to a variable named "region_type". 
- If the variables match the criteria for an enhancer, the value "enhancer" should be assigned to a variable named "region_type". 
- If the variables DO NOT match the criteria for a promoter and DO NOT match the criteria for an enhancer, the value "unknown" should be assigned to a variable named "region_type". 

Your code will be checked with different values of `rX_DHS`, `rX_dist_TSS`, `rX_H3K4me3`, and `rX_H3K27ac`, so make sure it will work for ALL possible situations.

Here is a reminder of the characteristics of promoter and enhancer regions:

|                   | Promoter region | Enhancer region      |
| ----------------- | --------------- | -------------------- |
| Distance from TSS | < 200 bp        | no specific distance |
| DHS               | TRUE            | TRUE                 |
| H3K4me3 level     | high            | low                  |
| H3K27ac level     | low             | high                 |

```{r Assignment.W2.3.c}
rX_DHS <- TRUE
rX_dist_TSS <- 398
rX_H3K4me3 <- "low"
rX_H3K27ac <- "high"
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.3.c.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### d.
(3 marks)
Use one or more conditional statements to write code that checks if region C is an enhancer and if so, checks if it is a *distal enhancer* (> 2000 bases from the TSS) or a *proximal enhancer* (<= 2000 bases from the TSS). 

- If the region is an enhancer, the value `TRUE` should be assigned to a variable named "is.enhancer".
- If the region is NOT an enhancer, the value `FALSE` should be assigned to a variable named "is.enhancer".
- If the region is an enhancer and the enhancer is distal, the value "distal" should be assigned to a variable named "enhancer.type".
- If the region is an enhancer and the enhancer is proximal, the value "proximal" should be assigned to a variable named "enhancer.type".

```{r Assignment.W2.3.d}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W2.3.d.R")
```
**IGNORE the above block: test results will not appear in assignments** 

---